DCMIP2016, Part 1: Models and Equation Sets

### 2016-08-18 contributions/GEM/GEM-DCMIP2016-Part1-Models.tex
Added contributions from modeling groups

@@

\subsection{FV$^3$} \label{sec:FV3Equations}

%%%GFDL start
FV$^3$ solves the fully-compressible shallow-atmosphere Euler equations using the vector-invariant equations for horizontal momentum on a mass-based Lagrangian vertical coordinate \citep{L2004MWR}. 
%%%GFDL end

%They can be written as:%
%
%\begin{align}
%\pdiff{\rho}{t} &= - \nabla \cdot (\rho \vb{u}), \\
%\pdiff{\vb{u}}{t} &= \left( f\vg{\hat k + \zeta} \right ) \times \vb{u} - \frac{1}{\rho}\nabla p |_z, \\
%\pdiff{\theta_v}{t} &= - \vb{u} \cdot \nabla \theta_v, \\
%\pdiff{w}{t} + \vb{u} \cdot \nabla w &= -\frac{1}{\rho}\pdiff{p}{z} - g,
%\end{align} in terms of Kinetic energy $K = \vb{u} \cdot \vb{u}$, relative vorticity $\vg{\zeta} = \nabla \times \vb{u}$, and local Coriolis parameter $f$. Here, gradient operators (except for that in the pressure-gradient term) are along Lagrangian surfaces.

@@

\subsection{Tracer transport}

@@

\subsection{Mass-based coordinates}

@@

{\color{red}[ALL] Include explicit diffusion and stabilization techniques that you have applied in the dynamical core here.}
%%%GFDL start

Explicit dissipation in FV$^3$ is applied separately to the divergence and to the horizontal fluxes in the governing equations. The D-grid discretization applies no direct implicit dissipation to the divergence, so divergence damping is an intrinsic part of the solver algorithm since otherwise there are no processes by which energy contained in the divergent modes is removed at the grid scale.  FV$^3$ has options for fourth-, sixth-, or eighth-order divergence damping; a second-order option is also available for use in idealized convergence tests, which can be applied in addition to the higher-order diffusion. The monotonicity constraint used in computing the fluxes in the  momentum, thermodynamic, and mass continuity equations is sufficient to damp and stabilize the non-divergent component of the flow. If additional damping is desired, or if the monotonicity constraint has been disabled, there is an option to apply hyperdiffusion to the fluxes in each of these equations, with the exception of the tracer transport, which always uses monotonic transport with no explicit diffusion. This dissipation is of the same order of accuracy as the divergence damping; however the nondimensional coefficient needed is much smaller, usually by at least 3--4 times, than the divergence damping. Both divergence damping and hyperdiffusion are applied along the Lagrangian surfaces and are re-computed every acoustic timestep; there is no explicit damping perpendicular to the Lagrangian surfaces. An option to convert kinetic energy lost to the hyperdiffusion to heat is available. FV$^3$ can also apply a horizontal Smagorinsky eddy diffusion if desired.

A substantial amount of wave absorption is provided by the flexible-lid (constant-pressure) upper boundary;  FV$^3$ also applies second-order diffusion to all fields, except the tracers, to create a sponge layer, typically comprising the top two layers of the domain, to damp other signals reaching the top of the domain. An energy-conserving Rayleigh damping, applied consistently to all three components of the winds, is also available, which is strongest in the top layer of the domain and becomes weaker with distance until reaching a runtime-specified cut-off pressure.

%%%GFDL end

@@

\subsection{Mass borrowing (positive definite preservation)}

@@

@@

\subsection{Global Environmental Multiscale (GEM) Model}

{\color{red}[LEE]}

@@

%% Since the Copernicus LaTeX package includes the BibTeX style file copernicus.bst,
%% authors experienced with BibTeX only have to include the following two lines:
%%
\bibliographystyle{copernicus}
\bibliography{DCMIP2016.bib,FV3.bib}
%%
%% URLs and DOIs can be entered in your BibTeX file as:
%%
%% URL = {http://www.xyz.org/~jones/idx_g.htm}
%% DOI = {10.5194/xyz}

>>>>

\subsection{Yin-Yang grid} \label{sec:GEM_model_grid}
In GEM model, we use spherical coordinates on the overset Yin-Yang grid.
The Yin-Yang grid  \citep{kageama2004yinyang}  has two grid components which are geometrically
identical (see Fig. \ref{fig:gem_yinyang}). They are combined to cover a spherical surface with partial
overlap on their borders. Each component is in fact a part of the
latitude-longitude grid and is defined in spherical
polar coordinates by
\begin{equation}\label{eq:gem_yinyang}
(-\frac{\pi}{4}-\delta_{\theta} \leq \theta \leq
\frac{\pi}{4}+\delta_{\theta})  \cap
(-\frac{3\pi}{4}-\delta_{\lambda} \leq \lambda \leq
\frac{3\pi}{4}+\delta_{\lambda}),
\end{equation}
where $\delta_{\lambda}, \delta_{\theta}$ are small buffers, which are
proportional to the respective grid-spacings, required for a minimum
overlap in the overset methodology. In our DCMIP experiments, we use 
$ \delta_{\theta}=2 $ degrees and
$ \delta_{\lambda}=3 \delta_{\theta} $.
%----------------------------------------------------
\begin{figure}[]
\includegraphics[height=6cm]{gem_yinyang.png}
\caption{ Yin-Yang grid }
\label{fig:gem_yinyang}
\end{figure}
%----------------------------------------------------

@@

\subsection{GEM} \label{sec:GEM_equations}
The equations of GEM model \citep{Girard2014} written out for a log-hydrostatic-pressure like coordinate $\zeta$  system co-rotating with the earth are:
\begin{eqnarray}
\frac{ {d \bf V}_{h} }{dt}+ f {\bf k} \times {\bf{V}_{h}} + R_d T_v\nabla
_\zeta \,  \ln\, p+ \left( 1 + \mu \right) \nabla
_\zeta {\phi} = 0,
\label{eq:gem_eq_1} \\
\frac{d w}{dt}- g \mu = 0,
\label{eq:gem_eq_2} \\
\frac{d}{dt} \ln \left( \pi \pdiff{\ln \pi}{\zeta}\right)
+ \nabla
_{ \zeta }\cdot {\bf V}_{{h}}+\frac{\partial
\dot{ \zeta }}{\partial
\zeta }  = 0,
\label{eq:gem_eq_3} \\
\frac{d \ln T_v}{dt} - \kappa \frac {d \ln p} {dt}=0,
\label{eq:gem_eq_4} \\
R_dT_v+  \frac{p}{\pi} \frac{\partial \phi}{\partial \ln \pi}=0,
\label{eq:gem_eq_5} \\
\frac{d \phi}{dt}-  g w  = 0,
\label{eq:gem_eq_6} \\
1+ \mu - \frac{p}{\pi}   \frac{\partial \ln p}{\partial \ln  \pi}=0,
\label{eq:gem_eq_7} \\
\ln \pi= \zeta+Bs ,
\label{eq:gem_eq_8}
\end{eqnarray}

\noindent with horizontal velocity ${\bf V}_h$, 
generalized vertical velocity $\dot{ \zeta }=\frac{d\zeta}{dt}$, 
vertical velocity $w$,
geopotential $\phi$, 
virtual temperature $T_v$, 
pressure $p$ and
hydrostatic pressure $\pi$. The variable 
$\mu$ is a measure of departure from hydrostatic balance, 
$s$ is related to the surface pressure $\pi_s$ and $B$ is a metric term. The last two variables are defined in (\ref{eq:gem_zeta}).
The equations contain a total time derivative $\frac{ {d } }{dt}$ that expresses the change in time while following the air parcel.

@@

\subsection{Tracer transport}
In GEM model, the transport equation for tracers uses the Lagrangian form:
\begin{align}
\diff{q}{t} &= 0.
\end{align}

\vfill
\vfill

@@

\subsection{Mass-based coordinates} \label{sec:GEM_zeta}
The vertical coordinate of GEM model named $ \zeta$ is of a log-hydrostatic-pressure type and is defined by:
\begin{align}\label{eq:gem_zeta}
 \ln \pi= \zeta+B(\zeta) s \, \, ; \, \, B(\zeta) =\left( \frac {\zeta-\zeta_T} {\zeta_s-\zeta_T}\right)^r \,; \, s= \ln ( \frac {\pi_s} {p_{o}}) \,; \, \, p_{o}=1000 \,\, \mbox{hPa},
\end{align}
with $\zeta_T= \ln \pi_T$, $\zeta_s= \ln {p_{o}}$ and where $r$ is a variable exponent providing added freedom for adjusting the thickness of model layers over high terrain.
Figure \ref{fig:gem_charney-phillips} shows the Charney-Phillips grid in GEM model, giving the position occupied by each variable in the vertical domain.
Horizontal velocity ${\bf V}_h$, geopotential $\phi$ and $q=\ln(p/\pi)$ are on momentum levels.
Temperature $T$, vertical velocity $w$,
generalized vertical velocity $\dot{ \zeta }$ and tracers are on thermodynamic levels.

%----------------------------------------------------
\begin{figure}[]
\includegraphics[height=10cm]{gem_charney-phillips.png}
\caption{ Charney-Phillips grid }
\label{fig:gem_charney-phillips}
\end{figure}
%----------------------------------------------------

@@

{\color{red}[ALL] Include explicit diffusion and stabilization techniques that you have applied in the dynamical core here.}

Via an application of (\ref{eq:scalar_viscosity_direct}),
scalar viscosity is employed for wind components and tracers. 
A vertical sponge layer, also via an application of (\ref{eq:scalar_viscosity_direct}),
is employed on wind components and $T_v$ with a vertical modulation on the 6 top levels 
and a maximum damping coefficient of $.380000 \times 10^{6}\ \mbox{m\textsuperscript{2}/s}$ at the top.
For stabilization purpose, 
the temporal discretization of GEM presented  
in section \ref{sec:GEM_temporal} uses an off-centering parameter $b^{A} = 0.6$.

@@

\subsection{Mass borrowing (positive definite preservation)}
In GEM model, shape-preserving solutions are computed for tracers by using the quasi-monotone semi-Lagrangian (QMSL) method \citep{Bermejo1992QMSL}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection { The 2 time level semi-Lagrangian implicit time discretization in GEM model} \label{sec:GEM_temporal}

Consider a frictionless adiabatic prognostic equation of the form:
\begin{equation} 
\frac {dF}{dt}+G=0,
\label{eq:gem_temporal_O}
\end{equation}
\noindent where $F$ represents one of the prognostic variables and $G$ represents the remaining terms, some of which are non-linear.
Based on the semi-Lagrangian scheme,
The equation (\ref {eq:gem_temporal_O}) is discretized as:
\begin{equation} 
\frac{F^A-F^{D}} {\Delta t}+ b^A G^A + (1-b^A) G^D = 0,
\label{eq:gem_temporal_D}
\end{equation}
with $A \, \mbox{(Arrival)} = (\vg{r},t)$, $D \, \mbox{(Departure)} = (\vg{r}-\Delta \vg{r},t-\Delta t)$ 
and $\vg{r}$ is the 3D grid position. 
The displacements $\Delta \vg{r}$ are obtained by the iterative process:
\begin{equation} 
\Delta  \vg{r}^{i} =  \left [ b^A \vg{V}(\vg{r},t) + (1-b^A) \vg{V}(\vg{r}-\Delta \vg{r}^{i-1},t-\Delta t) \right ] \Delta t,
\label{eq:gem_temporal_traj} 
\end{equation} 
where  $\vg{V}$ is the 3D wind vector.
From (\ref {eq:gem_temporal_D}), we regroup unknown terms on the left-hand side 
and known terms on the right-hand side, that is:
\begin{align} 
\frac{F^A} {\tau} + G^A =  \frac{F^{D}} {\tau}- \beta G ^{D}= R,
\label{eq:gem_temporal_AD}
\end{align}
\noindent where $\tau= \Delta t b^A$ and $\beta = (1-b^A)/b^A$. 
Cubic Lagrange interpolation is used for upstream evaluations $(F^{D}, G^{D})$.
Each left-hand side term $\frac{F^A} {\tau}+ G^A$ is then split into a linear part $L$ and a nonlinear residual part $N$:   
\begin{align} 
\frac{F^A} {\tau} + G^A = L + N = R. 
\label{eq:gem_temporal_LN}
\end{align}
The linearization is done around a constant in time, horizontally homogeneous, reference state. 
This results in the non-linear problem:
\begin{align}
L^{i}=R-N^{i-1},
\label{eq:gem_temporal_solve}
\end{align}
which is solved with two iterations.   
In each iteration, the
linear system (\ref{eq:gem_temporal_solve})  
is reduced to an Helmholtz problem 
for one composite variable.
It 
is solved 
with a direct solver
using the Schwarz-type domain decomposition method on Yin-Yang grid \citep{Qaddouri2008schwarz}.
The composite variable solution is used to update the prognostic variables at time $t$.  
Our procedure involves another iterative process linked to the estimation of $\vg{V}(\vg{r},t)$ in the displacement calculations (\ref{eq:gem_temporal_traj}). 
This results in redoing the sequence of operations (\ref{eq:gem_temporal_traj})-(\ref{eq:gem_temporal_solve}). 
At each time step, the cubic Lagrange interpolation is used to update the static halo region of both panels of the Yin-Yang grid 
\citep{Qaddouri2011operational}.

@@

\subsection{Global Environmental Multiscale (GEM) Model} \label{sec:GEM_core}

In GEM model, we use the Yin-Yang grid and horizontal discretization is done on an Arakawa C grid.
The vertical coordinate $\zeta$ is of a log-hydrostatic-pressure type and vertical discretization is based  
on the Charney-Phillips grid.
A 2 time level semi-Lagrangian implicit time discretization is implemented as described in section \ref{sec:GEM_temporal}. 
A scalar viscosity is employed for wind components and tracers 
via an application of (\ref{eq:scalar_viscosity_direct}). 
Viscosity operations are applied after the completion of the dynamic time step.

@@

%% Since the Copernicus LaTeX package includes the BibTeX style file copernicus.bst,
%% authors experienced with BibTeX only have to include the following two lines:
%%
\bibliographystyle{copernicus}
\bibliography{DCMIP2016.bib}
%%
%% URLs and DOIs can be entered in your BibTeX file as:
%%
%% URL = {http://www.xyz.org/~jones/idx_g.htm}
%% DOI = {10.5194/xyz}

### 2016-08-18 contributions/HOMME/HOMME-DCMIP2016-Part1-Models.tex
Added contributions from modeling groups

\subsection{FV$^3$} \label{sec:FV3Equations}

%%%GFDL start
FV$^3$ solves the fully-compressible shallow-atmosphere Euler equations using the vector-invariant equations for horizontal momentum on a mass-based Lagrangian vertical coordinate \citep{L2004MWR}. 
%%%GFDL end

%They can be written as:%
%
%\begin{align}
%\pdiff{\rho}{t} &= - \nabla \cdot (\rho \vb{u}), \\
%\pdiff{\vb{u}}{t} &= \left( f\vg{\hat k + \zeta} \right ) \times \vb{u} - \frac{1}{\rho}\nabla p |_z, \\
%\pdiff{\theta_v}{t} &= - \vb{u} \cdot \nabla \theta_v, \\
%\pdiff{w}{t} + \vb{u} \cdot \nabla w &= -\frac{1}{\rho}\pdiff{p}{z} - g,
%\end{align} in terms of Kinetic energy $K = \vb{u} \cdot \vb{u}$, relative vorticity $\vg{\zeta} = \nabla \times \vb{u}$, and local Coriolis parameter $f$. Here, gradient operators (except for that in the pressure-gradient term) are along Lagrangian surfaces.

@@

>>>>

@@

HOMME stands for the High Order Method Modeling Environment which provides the framework for the CAM-SE and ACME-Atmosphere dynamical cores along with several experimental versions  including a non-hydrostatic model. It was designed to be mass and energy conserving with nearly optimal parallel scalability at large core counts. The CAM-SE dynamical core is a hydrostatic model that partitions the globe horizontally using an unstructured grid of quadrilateral elements. These elements are arrange by default in a cube-sphere structure, but variable-resolutions grids with conforming edges may also be employed. Each quadrilateral extends in the radial direction to form a column using the shallow-atmosphere approximation $r\approx a$ where $a$ is the mean radius of the Earth. It employs a hybrid terrain-following / pressure coordinate $\eta \in (0,1]$ which transforms smoothly from a pure pressure coordinate at the top of the model to a terrain-following surface-pressure coordinate at the bottom. A conventional vector-invariant form for the moist primitive equations is employed with prognostic zonal and meridional wind components $u,v$, temperature $T$, surface pressure $p_s$, and tracer mixing ratios $q_i$. The PDEs are discretized using a split approximation, with a nodal 4th order spectral element discretization in the horizontal and the mimetic (mass and energy conserving) 2nd order finite difference discretization of \cite{simmons1981energy} in the vertical. Fields are co-located in the horizontal in the sense that they share the same 4th order basis functions. Lorenz staggering is employed in the vertical with (u,v,T,p) placed on layer midpoints while vertical fluxes $m \dot \eta$ are placed on layer interfaces.

A floating Lagrange formulation is used for tracer advection and optionally for the vertical dynamics as well, in which vertical levels are advected with the fluid. After several time-steps, the levels are remapped onto the fixed Eulerian $\eta$ grid in order prevent pressure surfaces from crossing. Several limiter options are available including a sign-preserving limiter and a monotone optimization base limiter described in \cite{guba2014optimization}. Diffusion is applied in the horizontal with a 4th order hyperviscosity operator, decomposed into two applications of the Laplacian operator in divergence / vorticity form. The physics, tracer, dynamics, and vertical remapping schemes are applied in a time split manner with adjustable sub-cycling ratios. Tracer advection is discretize in time with a 3-stage 2nd order strong stability preserving (SSP) scheme of Spitteri and Ruuth \cite{spiteri2002new}. The dynamics are also discretized using one of several explicit Runge-Kutta times schemes. Physics-dynamics coupling may be achieved using a single adjustment at the end of the physics timestep, through tendency update every dynamics step, or by a hybrid of the two.

In addition to the primitive equation model, HOMME also hosts an experimental nonhydrostatic dynamical core which retains many of the features of the traditional CAM-SE model. This model makes use of the shallow-atmosphere Euler equations of motion in terrain-following $\eta$ coordinates, as first described by \cite{laprise1992euler}. In its current incarnation, unstructured spectral elements are retained in the horizontal with matching spectral elements in the vertical. Prognostic variables include the 3d wind velocity components $u,v,w$, the surface pressure $p_s$, potential temperature $\theta$, and geopotential height $\phi$. Fields are current co-located in the vertical and explicit time-stepping is used. Planned improvements call for vertically implicit time stepping, and possible Charney-Phillips staggering. This model has been successfully applied to the dry DCMIP-2012 dynamical core tests including (vertical SE) tracer transport, orogoraphic waves, gravity waves, and the dry baroclinic instability. Moist dynamics, physics/dynamics coupling, and implicit time integration routines are currently under development.
